Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-06-19 pirouette_article.tex
Accept @Giappo's suggestions, simplify example code, progress in answering response letter

  \textbf{1. }
    Phylogenetic trees are currently routinely reconstructed from an alignment 
    of character sequences (usually nucleotide sequences). 
    Bayesian tools, such as MrBayes, RevBayes and BEAST2, 
    have gained much popularity over the last decade, 
    as they allow joint estimation of the posterior distribution of the 
    phylogenetic trees and the parameters of the underlying inference model.  
    An important ingredient of these Bayesian approaches is the species tree 
    prior.
    In principle, the Bayesian framework allows for comparing 
    different tree priors, which may elucidate the macroevolutionary 
    processes underlying the species tree.
    In practice, however, only macroevolutionary models that allow for fast computation of the prior probability are used.
    The question is how accurate the tree estimation is when the real macroevolutionary processes are substantially different from those assumed in the tree prior. \\
    \textbf{2. }
    Here we present \verb;pirouette;, 
    a free and open-source R package that assesses 
    the inference error made by Bayesian phylogenetics for a given macroevolutionary diversification model. \verb;pirouette; makes use of 
    BEAST2, but its philosophy applies to any Bayesian phylogenetic inference 
    tool. \\
  \textbf{3. }
    \iffalse
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples.
    \fi
    We describe \verb;pirouette;'s usage \new{providing full examples in which we interrogate a model for its power to describe another.
    }
    \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples and their 
    interpretation. \\
\end{abstract}

@@

\new{
  % 00002
  In this paper, we focus on BD models, instead of coalescent models [Explain
  these a bit],
  although these are just as important. 
}

@@

For example, one can add novel diversification models 
by writing a BEAST2 package that contains the likelihood 
formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Many diversification models (and their associated probability algorithms) 
have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
or diversity-dependent [\cite{etienne2012diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine}].
Other models treat speciation as a process that takes 
time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
or where diversification rate
depends on one or more traits [\cite{maddison2007estimating, fitzjohn2012diversitree}].
Despite such a rich theoretical landscape, only a few of these 
diversification models are available as tree priors in BEAST2
% 00005
\giovanni{
  PROPOSAL:
  Replace
  "Despite such a rich theoretical landscape, only a few of these diversification models are available as tree priors in BEAST2"
  with
  "Despite the fact that some models are already included as tree priors in BEAST2 (e.g. coalescent models, BD skyline, etc.) many others for which a likelihood formula is already available in literature, are not implemented yet"
}
\richel{
  This is awkward: most of these models are available as plug-ins,
  which was the whole idea when making BEAST2. This is a great moment
  to mention the tree models available in those BEAST2 packages
}
.

@@

\new{
    % 00007
    We choose to preserve the original topology to increase the similarity between the twin to the original tree. This works well in the cases of the birth-death or diversity-dependent models we consider in our examples.
    However this might not be suited for models in which branching times are strongly influenced by topology.
}

\new{
  % 00003
  The default option for the twin diversification model $p_T$ 
  is to use the standard BD model.
}
It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find 
the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a BD model) 
that maximize this likelihood applied 
to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. 
We simulate the new branching times using the TESS 
package [\cite{TESS, hohna2016tess}].

@@

# Save
library(ggplot2)
pir_outs <- list()
pir_outs[[1]] <- pir_out
pir_plots(pir_outs) +
  ggtitle("pirouette example 30") +
  ggsave("errors.png", width = 7, height = 7)

>>>>

  \textbf{1. }
    Phylogenetic trees are currently routinely reconstructed from an alignment 
    of character sequences (usually nucleotide sequences). 
    Bayesian tools, such as MrBayes, RevBayes and BEAST2, 
    have gained much popularity over the last decade, 
    as they allow joint estimation of the posterior distribution of the 
    phylogenetic trees and the parameters of the underlying inference model.  
    An important ingredient of these Bayesian approaches is the species tree 
    prior.
    In principle, the Bayesian framework allows for comparing 
    different tree priors, which may elucidate the macroevolutionary 
    processes underlying the species tree.
    In practice, however, only macroevolutionary models that allow for fast computation of the prior probability are used.
    The question is how accurate the tree estimation is when the real macroevolutionary processes are substantially different from those assumed in the tree prior. \\
    \textbf{2. }
    Here we present \verb;pirouette;, 
    a free and open-source R package that assesses 
    the inference error made by Bayesian phylogenetics for a given macroevolutionary diversification model. \verb;pirouette; makes use of 
    BEAST2, but its philosophy applies to any Bayesian phylogenetic inference 
    tool. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage \new{providing full examples in which 
    we interrogate a model for its power to describe another.
    \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples and their 
    interpretation. \\
\end{abstract}

@@

\new{
  % 00002
  In this paper, we focus on BD models, instead of coalescent models,
  although these are just as important. Coalescent models ... 
  [RJCB: Describe coalescent models]
}

@@

For example, one can add novel diversification models 
by writing a BEAST2 package that contains the likelihood 
formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Many diversification models (and their associated probability algorithms) 
have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
or diversity-dependent [\cite{etienne2012diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine}].
Other models treat speciation as a process that takes 
time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
or where diversification rate
depends on one or more traits [\cite{maddison2007estimating, fitzjohn2012diversitree}].
Despite such a rich theoretical landscape, only a few of these 
diversification models are available as tree priors in BEAST2
% 00005
\new{
  Despite the fact that some models are already included as tree priors 
  in BEAST2 (e.g. coalescent models, BD skyline, etc.) 
  many others for which a likelihood formula is already available 
  in literature, are not implemented yet.
}
\new{
  BEAST2 allows to extend its functionality by plug-ins, such
  as [RJCB: name some plug-ins with refs to literature].
}

@@

\new{
  % 00007
  We choose to preserve the original topology to 
  increase the similarity between the twin to the original tree. 
  This works well in the cases of the birth-death or diversity-dependent models 
  we consider in our examples.
  However, this might not be suited for models in which branching times 
  are strongly influenced by topology.
}

\new{
  % 00003
  The default option for the twin diversification model $p_T$ 
  is to use the standard BD model.
}
\new{
  % 00009
  \verb;pirouette; has a built-in function to use a Yule model as well.
  Additionally, a user can specify a function to generate a twin tree
  from any speciation model, such as, for example, a coalescent model.
}

It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find 
the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a BD model) 
that maximize this likelihood applied 
to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. 
We simulate the new branching times using the TESS 
package [\cite{TESS, hohna2016tess}].

@@

# Save
library(ggplot2)
pir_plot(pir_out) +
  ggtitle("pirouette example 30") +
  ggsave("errors.png", width = 7, height = 7)
